---
title: "A Brief Introduction to R"
author: Gavin L. Simpson
date: CSEE 2015 • May 20th 2015
fontsize: 10pt
classoption: "compress, aspectratio=169"
output:
  beamer_presentation:
    theme: m
    keep_tex: true
    highlight: tango
    slide_level: 2
    template: ./slides-template.tex
    fig_width: 6
    fig_height: 3.5
    fig_crop: false
    pandoc_args: [
      "--latex-engine=xelatex"
    ]
---

```{r setup-options, echo = FALSE, results = "hide", message = FALSE}
knitr::opts_chunk$set(comment=NA, fig.align = "center",
                      out.width = "0.7\\linewidth",
                      echo = TRUE, message = TRUE, warning = TRUE,
                      cache = TRUE)
knitr::knit_hooks$set(crop.plot = knitr::hook_pdfcrop)
```

```{r packages, echo = FALSE, results = "hide", message = FALSE, warning = FALSE}
library("ggplot2")
library("gapminder")
library("gridExtra")
```

## Why R?

Why use a complicated, command-line driven stats package like R?

- It's free!
- Widely used by statisticians for new statistical methods
- If something doesn't work the way you like, you can change it
- R is a programming language --- you can write your own functions
- R scripts, \alert{Sweave}, & \alert{knitr} for reproducible research
- Works everywhere (Windows, OS X, Linux, ...)
- R was designed for expressive data analysis

# R Basics

## R Basics

Start \alert{RStudio} --- take a look around

R will be running in either your home directory or where is was installed

Set the working directory to folder containing your analysis

## Getting Help

R comes with a lot of documentation

To get help on functions or concepts within R, use the `"?"` operator

For help on the `getwd()` function use: `?getwd`

Function `help.search("foo")` will search through all packages installed for help pages with `foo` in them

R-Help mailing list: \url{http://www.r-project.org/mail.html}

## Getting Help: Task Views

\url{cran.r-project.org/web/views/}

\begin{figure}
\includegraphics[height=6cm,keepaspectratio=true]{figs/cran-task-views}
\end{figure}

## Getting Help: Stack Overflow

\url{stackoverflow.com/questions/tagged/r}

\begin{figure}
\includegraphics[height=6cm,keepaspectratio=true]{figs/stackoverflow-screen-grab}
\end{figure}

## Working with R

Type commands at the prompt `>` --- R evaluates them when you hit `Return`

If a line is *not* syntactically complete, the prompt changes to `+`

Create an object by assigning something to it

\scriptsize
```{r assignment}
radius <- 5
pi * radius^2
```
\normalsize

If we don't assign, R prints a *representation* of the object

## Packages

R comes with a basic set of functionality plus some \alert{recommended} packages

Additional functionality added via \alert{packages} from **CRAN**, **github**, **Bioconductor**, **drat** repos

\scriptsize
```{r load-packages-example, eval = FALSE}
install.packages(c("gapminder", "ggplot2"))
library("gapminder")
library("ggplot2")
```
\normalsize

## Reading data into R

\scriptsize
```{r read-gapminder}
gap <- system.file("gapminder.tsv", package = "gapminder")
gapminder <- read.delim(gap)
head(gapminder)
```
\normalsize

## This is where we are heading

\scriptsize
```{r gapminder-example, out.width = "0.6\\linewidth"}
ggplot(subset(gapminder, continent != "Oceania"),
       aes(x = year, y = lifeExp, group = country, color = country)) +
  geom_line(show_guide = FALSE) + facet_wrap(~ continent) +
  scale_color_manual(values = country_colors) +
  theme_bw()
```
\normalsize

## R Objects

What kind of object is `gapminder`?

\scriptsize
```{r str-gapminder}
str(gapminder)
class(gapminder)
```
\normalsize

## Data frames

A \alert{data frame} is R's version of an Excel spreadsheet

Columns are variables

Rows are observations

Different **types** of data in columns

Each column (\alert{component}) is of the same length

Is a special case of a list

## Subsetting

Access the columns of a data frame using `[`, `[[` or `$`

`$` is simple:

Access just a single variable

\scriptsize
```{r subset-dollar}
head(gapminder$country)
```
\normalsize

Uses the name of required variable

Partial matching

\scriptsize
```{r subset-dollar-2}
head(gapminder$cou)
```
\normalsize

## Subsetting

`[[` is a little more flexible:

Access just a single variable

Use the name of required variable

\scriptsize
```{r subset-double-bracket}
head(gapminder[["continent"]])
```
\normalsize

Or select the *n*th component

\scriptsize
```{r subset-double-bracket-2}
head(gapminder[[2]])
```
\normalsize

Partial matching optional --- `gapminder[["cont", exact = FALSE]]`

## Subsetting

`[` is a even more flexible:

Access one or more variables

Use the name(s) of required variable

\scriptsize
```{r subset-single-bracket}
head(gapminder[c("country", "continent")])
```
\normalsize

## subsetting

Or select the *n*th component(s)

\scriptsize
```{r subset-single-bracket-2}
head(gapminder[1:2])
```
\normalsize

## subsetting

Or we can index by rows and columns: `[rows, cols, other_args]`

\scriptsize
```{r subset-single-bracket-3}
head(gapminder[1:4, c(1,3)])
```
\normalsize

## subsetting

Leaving the row or column identifier *blank* means **"give me all of the rows (columns)"**

\scriptsize
```{r subset-single-bracket-4}
head(gapminder[1:4, ])                  # all columns, rows 1--4
```
\normalsize

\scriptsize
```{r subset-single-bracket-5}
head(gapminder[, 1:3], 3)               # all rows, columns 1--3
```
\normalsize

## subsetting

Empty dimensions get **dropped** if you select a single column

\scriptsize
```{r subset-single-bracket-6}
head(gapminder[, 2], 3)                 # just column 2
```
\normalsize

Preserve dimensions using `drop = FALSE`

\scriptsize
```{r subset-single-bracket-7}
head(gapminder[, 2, drop = FALSE], 3)   # all rows, columns 1--3
```
\normalsize

## Subsetting

Can use a range of \alert{index} types

- Numeric values select the `n`th elements
- Negative numeric values select all but those elements
- Character values select elements by name (possibly with partial matching)
- Logical values select (`TRUE`) & deselect (`FALSE`) elements
- Logical indices are \alert{recycled} to the correct length

\scriptsize
```{r subsetting-types}
(1:10)[c(FALSE, TRUE)]
(1:9)[c(FALSE, TRUE)]                   # no warning
```
\normalsize

## Vectors

What are the columns of `gapminder`?

\scriptsize
```{r str-gapminder-2}
str(gapminder)
```
\normalsize

Each component is a vector, of which there are several types: `numeric`, `character`, `logical`, `factor`, `integer`

## Vectors --- numeric & integer

R normally stores numeric data as *doubles* (decimal values)

There is an *integer* type too

\scriptsize
```{r numeric-vectors-1}
class(gapminder$lifeExp)
class(gapminder$year)
```
\normalsize

## Vectors --- numeric & integer

Create numeric vectors using `c()` or `:`

\scriptsize
```{r numeric-vectors-2}
c(1,3,5,7,9)
1:10
```
\normalsize

`x:y` is shorthand for `seq(from = x, to = y, by = 1)`

## Vectors --- character

Character vectors contain text (strings)

\scriptsize
```{r character-vectors-1}
c("foo","bar")
```
\normalsize

Quote each string using single or double quotes

## Vectors --- logical

Logical vectors are vectors of `TRUE` or `FALSE` values

\scriptsize
```{r logical-vectors-1}
c(TRUE, TRUE, FALSE)
```
\normalsize

- `FALSE` is 0
- `TRUE` is anything else, but is coerced to `1`

\scriptsize
```{r logical-vectors-2}
as.numeric(c(TRUE, TRUE, FALSE))
```
\normalsize

## Vectors --- factors

Factors are a special kind of vector

- stored internally as a vector of *codes*
- the codes index a set of \alert{levels} or categories, which can be numeric of character

\scriptsize
```{r factors-1}
f <- factor(c("Male","Female","Male"))
levels(f)
f <- factor(c(1,2,5,5,2,1))
as.numeric(f)                           # WRONG! Gets internal codes
as.numeric(as.character(f))             # RIGHT! correct coercion
```
\normalsize

## Sequences & patterned vectors

Sequences and patterned vectors are very useful in some circumstances

\scriptsize
```{r seqs-patterns}
seq(from = 1, to = 10, by = 2)
1:5
rep(1:3, each = 2)
rep(1:3, times = 3:1)
```
\normalsize

# Functions

## Functions

Pretty much everything in R is either a \alert{function} or the result of a call to one

Called with following format: `fun_name(arg1 = value1, arg2 = value2)`

\scriptsize
```{r functions-1}
rnorm(10)
args(rnorm)
rnorm(10, mean = 2, sd = 4)
```

## Functions

Can use \alert{positional} matching for argument names, but don't except for the first

\scriptsize
```{r functions-2}
rnorm(10, 2, 4)                         # What the heck does this do?
```
\normalsize

If you name arguments can be in any order (can be partial names)

\scriptsize
```{r functions-3}
rnorm(sd = 4, mean = 2, n = 10)
```
\normalsize

## Functions

You can write your own functions using the `function()` function

\scriptsize
```{r functions-4}
foo <- function(x) {                    # foo() squares it's input
    x * x                               # last statement determines return value
}
class(foo)
foo(10)
```
\normalsize

# Split-apply-combine

## Split-apply-combine

The \alert{split-apply-combine} model is a common type of data analysis

- \alert{split} data into chunks based on one or more factors
- \alert{apply} a function to each chunk
- \alert{combine} the outputs of applying the function to each chunk

Several R packages provide consistent and efficient implementations of the split-apply-combine model

- **plyr**, **dply**, **data.table**

But base R has useful functions too

- `aggregate()`, `split()` + `apply()`-family + `c|rbind()`

## Split-apply-combine

`aggregate` applies `FUN` to a vector, split up by one or more factors:

\scriptsize
```{r split-apply-combine-1}
aggregate(pop ~ continent, data = gapminder, FUN = median)
```
\normalsize

## Split-apply-combine

Can do this by hand too

\scriptsize
```{r split-apply-combine-2}
with(gapminder, sapply(split(pop, f = continent), FUN = median)) 
```
\normalsize

## Split-apply-combine --- apply family

The `apply` family provides very general approaches to applying function to aspects of data

- `apply` applies a function to the `MARGIN`s of a matrix, array, or data frame
- `sapply` applies the function to components of a list or data frame & **simplifies** if possible
- `lapply` applies the function to components of a list or data frame & returns a list
- `tapply` applies the function to chunks of data created by splitting on a factor
- `mapply`, `vapply()`, `rapply()` are specialist alternatives

## Split-apply-combine --- apply family

\scriptsize
```{r split-apply-combine-3}
apply(gapminder[, 4:5], 2, FUN = median)
tapply(gapminder$pop, gapminder$continent, FUN = median)
```
\normalsize

## Split-apply-combine --- apply family

\scriptsize
```{r split-apply-combine-4}
with(gapminder, lapply(split(pop, f = continent), FUN = median))
```

# Modelling

# Plotting

## Plotting

Your R installation comes with two main plotting toolboxes

- base graphics
- grid graphics

Grid graphics is extremely flexible but that comes at a cost of complexity

Two high-level interfaces to grid provides extensive plotting capabilities

- **lattice**, which comes with R
- **ggplot2**, which needs to be installed from CRAN

## Base graphics

These are the standard types of plots available in & produced by R & add-on packages

The main function is `plot()`, with `points()`, `lines()`, `text()`, `segments()`, `polygons()`, etc acting as lower-level elements

The look and feel of the plots is essentially controlled via \alert{graphical parameters} --- `?par`

Other high-level functions provide to access to the main plot types --- `boxplot()`, `hist()`, `stripchart()`, `barchart()`

## Base graphics

\scriptsize
```{r base-plot-1, crop.plot = TRUE}
x <- rnorm(20)
y <- rnorm(20)
plot(x, y, pch = 1, col = "navyblue", cex = 1.2, type = "p")
```
\normalsize

## Base graphics

\scriptsize
```{r base-plot-2, crop.plot = TRUE, eval = FALSE}
plot(x, y, lty = "dashed", col = "navyblue", cex = 1.2, type = "l")
plot(x, y, pch = 2, col = "navyblue", cex = 1.2, type = "o")
```
\normalsize

```{r base-plot-3, crop.plot = TRUE, echo = FALSE, out.width = "0.85\\linewidth"}
layout(matrix(1:2, ncol = 2))
plot(x, y, lty = "dashed", col = "navyblue", cex = 1.2, type = "l")
plot(x, y, pch = 2, col = "navyblue", cex = 1.2, type = "o")
layout(1)
```

## Base graphics

\scriptsize
```{r base-plot-4, crop.plot = TRUE, out.width = "0.55\\linewidth", fig.show = "hold"}
op <- par(mar = c(5,4,5,4) + 0.1)       # alter margins
plot(x, y, pch = 1, col = "navyblue", cex = 1.2, type = "p", ann = FALSE, axes = FALSE)
axis(side = 1); axis(side = 2)          # add axis, can be customised
axis(side = 3); axis(side = 4)
box()                                   # draw the box round the plot
title(main = "My plot", xlab = "x-axis", ylab = "my y-axis label")
par(op)                                 # reset plotting parameters
```
\normalsize

## GGplot

Base graphics are serviceable but require a lot of cruft code to go beyond basic plots --- encoding size, colour, etc using data

This is where **lattice** and **ggplot2** graphics come in

These are high-level plotting toolboxes that provide interfaces espousing Trellis Graphics and The Grammar of Graphics ideas, both built on top of **grid**

These are *not* general purpose graphics toolkits --- need to follow the ideas & theory behind the respective paradigm

If you want general-purpose, you need to use base graphics or grid

Can't (easily) mix base and grid graphics

## ggplot

\alert{ggplot2} is an implementation of Leland Wilkinson's Grammar of Graphics (hence `gg` in the name)

Three key components of a **ggplot2** plot

- **data** --- the data must be in the form of a data frame
- **aesthetics** --- how should data be represented on the plot
    * essentially **mappings** from variables to coordinates, size, colour, shape, transparency
- **geometries** --- how to physically draw the data & mappings

ggplot graphics consist of zero or more layers

Additionally, **stats** transform variables, **scales** control axis scaling & legends, **themes** control overall look & feel, **facets** split data into panels

## ggplot --- a basic plot

\scriptsize
```{r ggplot-1, fig.width = 7}
library("ggplot2")             # load the package
plt <- ggplot(gapminder, mapping = aes(x = year, y = pop, colour = continent, group = country)) +
         geom_line()           # add a layer with a line geometry
plt                            # Have to print the object to draw plot
```
\normalsize

## ggplot --- faceting

\scriptsize
```{r ggplot-2, fig.width = 9, out.width = "0.9\\linewidth"}
plt + facet_wrap(~ continent)  # facet by continent
```
\normalsize

## ggplot --- stats

\scriptsize
```{r ggplot-3, fig.width = 9, out.width = "0.9\\linewidth"}
ggplot(gapminder, mapping = aes(x = continent, y = lifeExp)) +
    geom_boxplot()                      # has a default stat "boxplot"
```
\normalsize

## ggplot --- stats

\scriptsize
```{r ggplot-4, fig.width = 9, out.width = "0.9\\linewidth"}
ggplot(gapminder, mapping = aes(x = lifeExp)) +
    geom_histogram()                      # has a default stat "bin"
```
\normalsize

## ggplot --- stats

\scriptsize
```{r ggplot-5, fig.width = 9, out.width = "0.9\\linewidth"}
ggplot(gapminder, mapping = aes(x = lifeExp)) +
    geom_line(stat = "density")         # change the stat
```
\normalsize

## ggplot --- grouping & smooths

\scriptsize
```{r ggplot-6, fig.width = 7, out.width = "0.7\\linewidth"}
(p2 <- ggplot(subset(gapminder, year == 2002 & continent != "Oceania"),
             aes(x = gdpPercap, y = lifeExp, colour = continent, group = continent)) +
        geom_point() + geom_smooth(method = "lm"))
```
\normalsize

## ggplot --- Themes

\scriptsize
```{r ggplot-7, fig.width = 7, out.width = "0.7\\linewidth"}
p2 + theme(legend.position = "top")     # move the legend, see ?theme
```
\normalsize

## ggplot --- Present Themes

\scriptsize
```{r ggplot-8, fig.width = 7, out.width = "0.7\\linewidth"}
p2 + theme_bw()                         # a simple theme
```
\normalsize

## ggplot --- Exporting plots

ggplot plots can be exported to disk using `ggsave()`

- If the image is on screen (last plot)
    ```{r ggsave-1, eval = FALSE}
    ggsave(file = "filename.pdf")
    ```
- If the plot is saved as an object
    ```{r ggsave-2, eval = FALSE}
    ggsave(p2, file = "filename.pdf")
    ```
- Specify the size
    ```{r ggsave-3, eval = FALSE}
    ggsave(file = "filename.pdf", width = 6, height = 4)
    ```
- Change the file type by modifying the extension
    ```{r ggsave-4, eval = FALSE}
    ggsave(file = "filename.png")
    ```

## Re-use

Unless indicated otherwise, this slide deck is licensed under a [Creative Commons Attribution 4.0 International License](http://creativecommons.org/licenses/by/4.0/).

\begin{center}
  \ccby
\end{center}
